---
title: "Complete MEA Analysis Example"
author: "Alex Tudoras"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete MEA Analysis Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library("NOVA")
```

# MEA Data Analysis Pipeline

This vignette demonstrates a complete NOVA workflow using two WT neuron
experiments (MEA012, MEA013) treated with a panel of neuronal agonists
(PBS, KCl, DHPG, Gabazine, AMPA, KA). All wells are wild-type; there is
no genotype split in this dataset.

The dataset ships with the NOVA package in `Example/MEA Neuronal Agonists/`.

## Package Installation and Loading

```{r package_setup}
# Install NOVA from GitHub if needed:
# devtools::install_github("atudoras/nova")
library(NOVA)
```

## 1. Discover Data Structure

```{r discover}
main_directory <- "MEA Neuronal Agonists"

discovery_results <- discover_mea_structure(main_directory)

cat("Experiments found:", discovery_results$experiment_count, "\n")
cat("Experiments     :", paste(names(discovery_results$experiments), collapse = ", "), "\n")
cat("Timepoints      :", paste(discovery_results$all_timepoints, collapse = ", "), "\n")
```

## 2. Process Data

Both experiments are WT-only, so `Genotype` is not used as a grouping
variable. The union of timepoints from MEA012 and MEA013 is used; missing
entries for a given experiment/timepoint combination are handled internally.

```{r process}
all_data <- process_mea_flexible(
  main_dir             = main_directory,
  selected_experiments = NULL,   # use all discovered experiments
  selected_timepoints  = c("baseline", "0min", "15min", "30min", "45min",
                           "1h", "1h15", "1h30", "1h45", "2h"),
  grouping_variables   = c("Experiment", "Treatment", "Well"),
  baseline_timepoint   = "baseline",
  verbose              = TRUE
)

raw_data        <- all_data$raw_data
normalized_data <- all_data$normalized_data

cat("\nRaw data dimensions      :", dim(raw_data), "\n")
cat("Normalized data dimensions:", dim(normalized_data), "\n")
```

## 3. Summary Statistics

```{r summary}
library(dplyr)

raw_data %>%
  dplyr::group_by(Experiment, Timepoint) %>%
  dplyr::summarise(
    n_wells     = dplyr::n_distinct(Well),
    n_variables = dplyr::n_distinct(Variable),
    mean_value  = round(mean(Value, na.rm = TRUE), 3),
    .groups     = "drop"
  ) %>%
  print()
```

## 4. PCA Analysis

```{r pca}
pca_output <- pca_analysis_enhanced(processing_result = all_data)
```

## 5. PCA Plots – Color by Treatment

All wells are WT, so we color by `Treatment` only (no shape mapping for
genotype).

```{r pca_plots}
plot_results <- pca_plots_enhanced(
  pca_output      = pca_output,
  output_dir      = file.path(main_directory, "MEA_Plots"),
  color_variable  = "Treatment",
  shape_variable  = NULL,
  pannels_var     = NULL,
  components      = c(1, 2),
  plot_width      = 12,
  plot_height     = 10,
  experiment_name = "WT_Agonists",
  verbose         = TRUE
)
```

## 6. Variable Importance

```{r importance}
importance_results <- analyze_pca_variable_importance_general(
  pca_result   = pca_output,
  output_dir   = file.path(main_directory, "MEA_Plots"),
  color_scheme = "viridis",
  top_n        = 20
)
```

## 7. PCA Trajectories

```{r trajectories}
traj_results <- plot_pca_trajectories_general(
  pca_output,
  timepoint_order     = c("baseline", "0min", "15min", "30min", "45min",
                           "1h", "1h15", "1h30", "1h45", "2h"),
  individual_var      = "Well",
  trajectory_grouping = "Treatment",
  smooth_lines        = TRUE,
  output_dir          = file.path(main_directory, "MEA_Plots")
)
```

## 8. Heatmaps

```{r heatmaps}
heatmaps <- create_mea_heatmaps_enhanced(
  processing_result = all_data,
  grouping_columns  = "Treatment",
  output_dir        = file.path(main_directory, "MEA_Plots"),
  verbose           = TRUE
)
```

## Save Results

```{r save}
save(all_data, pca_output,
     file = file.path(main_directory, "nova_results_WT_agonists.RData"))
cat("Results saved.\n")
```
