---
title: "A generic NOVA pass separates a DNT training set's actives from its negative"
subtitle: "EPA/Shafer multi-well MEA network formation assay (Brown et al. 2016) — exploratory reanalysis"
date: today
params:
  study_dir: "."
  nova_root: "../.."
  findings_csv: "outputs/discovery/findings.csv"
  data_rds: "outputs/discovery/discovery.rds"
  source_md: "SOURCE.md"
  group_var: "Treatment"
  min_wells: 4
  n_top: 6
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    theme: cosmo
    fig-width: 8
    fig-height: 5
  typst:
    papersize: a4
    margin: {x: 1.8cm, y: 1.8cm}
    fig-format: png
    fig-dpi: 200
    fig-width: 6.5
    fig-height: 3.4
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({ library(dplyr); library(ggplot2); library(knitr) })

# NOVA is loaded from this repo's source, never the installed library: the installed
# copy here is 0.1.1 and predates the trajectory layer, so a report built against it
# would showcase different code than the repo it ships in.
suppressMessages(devtools::load_all(params$nova_root, quiet = TRUE))
stopifnot("nova_trajectory_summary" %in% getNamespaceExports("NOVA"))

findings <- read.csv(file.path(params$study_dir, params$findings_csv), stringsAsFactors = FALSE)
scan     <- readRDS(file.path(params$study_dir, params$data_rds))

# --- the source paper's designations, quoted, not remembered -------------------
# Brown et al. 2016, Toxicol Sci 154(1):126-139, doi:10.1093/toxsci/kfw147, abstract:
#   "All compounds except acetaminophen (<= 30 uM) caused concentration-related
#    effects on one or more of these parameters."
#   "Domoic acid and sodium orthovanadate altered several of these parameters in the
#    absence of cytotoxicity."
# Five actives, one negative -- and among the actives, two act without killing cells.
# See SOURCE.md. Spelling follows the data file, which misspells Bis-1.
DESIGNATION <- tibble::tribble(
  ~compound,                ~label,                  ~cytotox_free,
  "Acetaminophen",          "negative",              NA,
  "Domoic Acid",            "active",                TRUE,
  "Sodium Orthovanadate",   "active",                TRUE,
  "Bisindolymaleimide 1",   "active",                FALSE,
  "Mevastatin",             "active",                FALSE,
  "Loperamide",             "active",                FALSE,
  "Glyphosate",             "not in training set",   NA
)
ACTIVES  <- DESIGNATION$compound[DESIGNATION$label == "active"]
NEGATIVE <- DESIGNATION$compound[DESIGNATION$label == "negative"]
CYTOTOX  <- DESIGNATION$compound[which(DESIGNATION$cytotox_free == FALSE)]

moves <- findings[findings$kind == "move", ]
strip_dose <- function(x) sub(" \\(vehicle\\)$", "", sub(" [0-9.]+ uM$", "", x))
moves$compound <- strip_dose(moves$group)

# --- the one curation rule ----------------------------------------------------
# A median computed on one well is not a median. Ranking on raw effect size cannot
# tell three wells from thirty, so single-well conditions rise on the strength of one
# lucky (or one dying) well. This is a validity filter, not a change to the ranking:
# what survives is still ordered by raw effect size.
kept    <- moves[moves$n_wells >= params$min_wells, ]
dropped <- moves[moves$n_wells <  params$min_wells, ]
top     <- head(kept, params$n_top)

# Headline computed on BOTH the raw ranking and the filtered one: a claim that only
# survives one of them is a claim about the filter, not about the assay.
top25_raw  <- head(moves, 25)
top25_kept <- head(kept, 25)
pct <- function(x) sum(x)

# --- what the >0 filter removed, quantified -----------------------------------
# log2(0) is -Inf, so fully-silenced wells cannot enter a log-ratio ranking. Those
# are the largest possible effects, so their absence has to be stated, not assumed.
d_all <- scan$data
silenced <- d_all %>% filter(.data$Normalized_Value == 0)
lost_conditions <- setdiff(unique(d_all$Treatment), unique(moves$group))
lost_conditions <- setdiff(lost_conditions, grep("vehicle", lost_conditions, value = TRUE))
```

## Headline

```{r headline, results="asis"}
n_act_raw  <- sum(top25_raw$compound  %in% ACTIVES)
n_neg_raw  <- sum(top25_raw$compound  == NEGATIVE)
n_act_kept <- sum(top25_kept$compound %in% ACTIVES)
n_neg_kept <- sum(top25_kept$compound == NEGATIVE)
n_down_raw <- sum(top25_raw$direction == "down")

cat(sprintf(
  paste0("NOVA's standard pipeline was run over this dataset with no compound-specific ",
         "tuning: normalise to same-plate vehicle controls, PCA, trajectory summary, rank ",
         "by effect size. Brown et al. report that every compound in this training set ",
         "except acetaminophen produced concentration-related effects.\n\n"),
  ""))
cat(sprintf(
  paste0("**All %d of the 25 largest changes are among the paper's five actives, and ",
         "acetaminophen appears %d times.** The result does not depend on the curation ",
         "rule below: on the filtered ranking it is %d of 25 actives and %d acetaminophen. ",
         "%d of the raw top 25 are decreases relative to vehicle.\n\n"),
  n_act_raw, n_neg_raw, n_act_kept, n_neg_kept, n_down_raw))
cat(sprintf(
  # ASCII only inside R-emitted strings: text written by cat() into a
  # results="asis" chunk does not survive typst's encoding path, and an em-dash
  # renders literally as <U+2014>. Markdown prose in this file is unaffected.
  paste0("This is a statement about **prediction, not neurotoxicity**, and a weaker one ",
         "than it first looks (see the caveats): %d of the raw top 25 are compounds the ",
         "paper reports as cytotoxic, and the viability data that would separate ",
         "\"changed the network\" from \"killed the cells\" is not in this file.\n"),
  sum(top25_raw$compound %in% CYTOTOX)))
```

## Hero figure — the negative is flat, the actives are not

```{r hero, fig.cap="Synchrony index as a fraction of same-plate vehicle controls at DIV 7, for every compound with at least 4 replicate wells at a given dose. Dashed line = vehicle. Points are the condition median across replicate wells, counted by plate and well ID. Sodium orthovanadate falls monotonically to ~1% of vehicle. Acetaminophen -- the paper's only negative -- shows no dose trend, though it sits above vehicle throughout. Mevastatin appears as two points only: the n>=4 rule removes its higher doses, so this panel understates it."}
sync <- d_all %>%
  filter(.data$Variable == "Synchrony Index (r)", .data$Timepoint == "DIV7",
         !.data$Is_Control, is.finite(.data$Normalized_Value),
         .data$Normalized_Value > 0) %>%
  # Count wells by plate AND ID. n_distinct(Well) pools A1 across every plate into
  # one replicate: here that undercounts 28 of 32 conditions, 65 wells where there
  # are 142.
  mutate(.unit = NOVA::nova_unit_id(.)) %>%
  group_by(.data$Compound, .data$Dose) %>%
  summarise(fc = stats::median(.data$Normalized_Value),
            n = dplyr::n_distinct(.data$.unit), .groups = "drop") %>%
  filter(.data$n >= params$min_wells)

sync$Designation <- DESIGNATION$label[match(sync$Compound, DESIGNATION$compound)]
pal <- stats::setNames(NOVA::nova_palette(dplyr::n_distinct(sync$Compound)),
                       sort(unique(sync$Compound)))

hero <- ggplot(sync, aes(.data$Dose, .data$fc, colour = .data$Compound)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_line(aes(linetype = .data$Designation), linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = pal, name = NULL) +
  scale_linetype_manual(values = c(active = "solid", negative = "22"), name = NULL) +
  labs(x = "Dose (uM, log scale)", y = "Synchrony / vehicle control (log scale)",
       caption = "Lines join doses of one compound; no dose-response model was fitted.") +
  NOVA::nova_theme(base_size = 11)

# Also written to disk so the repo's README can show the result without anyone
# clicking anything -- GitHub renders neither a Quarto HTML notebook nor, as it
# turns out, this PDF. Exported from the same object the report plots, so the
# figure in the README cannot drift away from the figure in the study.
if (knitr::is_html_output()) {
  ggsave(file.path(params$study_dir, "hero.png"), hero,
         width = 7.5, height = 4, dpi = 140, bg = "white")
}
hero
```

## The largest changes that more than one well agrees on

```{r small-multiples, fig.height=3.6, fig.cap="Top changes with at least 4 replicate wells, selected by effect-size rank from findings.csv rather than by eye. Bars are median log2 fold-change against same-plate vehicle controls -- not against a baseline timepoint. Well counts are printed on each bar."}
if (nrow(top)) {
  t <- top
  t$label  <- paste0(t$metric, "\n", t$group, " @ ", t$timepoint)
  t$signed <- ifelse(t$direction == "down", -t$effect, t$effect)
  t$label  <- factor(t$label, levels = t$label[order(t$signed)])
  ggplot(t, aes(.data$label, .data$signed, fill = .data$compound)) +
    geom_col(width = 0.68) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text(aes(label = paste0("n=", .data$n_wells)),
              hjust = -0.25, size = 2.9, colour = "grey25") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
    scale_fill_manual(values = NOVA::nova_palette(dplyr::n_distinct(t$compound)), name = NULL) +
    coord_flip() +
    labs(x = NULL, y = "median log2 (fraction of vehicle control)") +
    NOVA::nova_theme(base_size = 9)
}
```

```{r ranked-table}
if (nrow(top)) {
  top %>%
    transmute(Condition = .data$group, Metric = .data$metric, DIV = .data$timepoint,
              `vs control` = sprintf("%.2fx %s", 2^.data$effect, .data$direction),
              Wells = .data$n_wells, Cultures = .data$n_experiments) %>%
    knitr::kable(row.names = FALSE, caption = sprintf(
      "Top %d changes with >= %d wells, by effect size. 'Cultures' counts distinct (date, plate) experiments, of which there are 6 across 4 physical plates.",
      nrow(top), params$min_wells))
}
```

::: {.content-visible when-format="html"}

## What the trajectory layer says, and why it misleads here

```{r describe, results="asis"}
txt <- utils::capture.output(NOVA::nova_describe(scan$trajectory))
cat("```\n"); cat(txt, sep = "\n"); cat("\n```\n")
```

Read this one carefully. `nova_trajectory_summary()` measures how far a condition travels
through state space **between DIV 7 and DIV 12** — not how different it is from control.
A compound that silences the network by DIV 7 and keeps it silent has a large effect and
almost no trajectory, so the largest movers here are *low* doses, whose effect is still
developing across the window. That is the metric behaving exactly as documented, answering
a question this design does not ask.

The same applies to group separation: several fully-silenced conditions land on identical
centroids, because everything at zero looks alike.

:::

## Caveats

::: {.callout-warning appearance="simple"}
**Exploratory. Nothing here is a tested claim.**

- **The ranking is dominated by compounds the source reports as cytotoxic.**
  `r sum(top25_raw$compound %in% CYTOTOX)` of the raw top 25 are Bis-1, mevastatin or
  loperamide. Brown et al. single out **domoic acid and sodium orthovanadate** as the two
  that "altered several of these parameters in the absence of cytotoxicity" — the
  implication being that the others' effects are at least partly cell death. The Cell
  Titer Blue viability data is **not in this file**, so this analysis cannot separate
  "changed the network" from "killed the cells". "Recovers the training set's actives" is
  therefore partly "recovers cytotoxicity".
- **Fully-silenced wells are absent from the ranking.** A log-ratio cannot represent zero,
  so `r nrow(silenced)` observations with a value of exactly 0 vs control are excluded —
  and those are the *largest possible* effects. `r length(lost_conditions)` conditions
  disappear entirely as a result (`r paste(lost_conditions, collapse = "; ")`), which are
  top doses of three actives. The ranking therefore systematically under-represents
  complete network silencing. This is also why the trajectory section counts 45 conditions
  while the ranking sees `r dplyr::n_distinct(moves$group)`.
- **Selection inflates the top of the list.** `r nrow(moves)` condition × metric ×
  timepoint comparisons were scanned and sorted by effect size; sorting a large family of
  noisy estimates puts the luckiest first. No correction is applied because nothing is a
  hypothesis test.
- **`r nrow(dropped)` of `r nrow(moves)` comparisons rest on fewer than
  `r params$min_wells` wells** and are excluded from the figures and table. Raw effect
  size cannot distinguish a large change in three wells from one in thirty: unfiltered,
  `r sum(top25_raw$n_wells == 1)` of the top 25 rest on a **single well**. The largest
  single-well effect is `r sprintf("%.0fx", 2^max(top25_raw$effect[top25_raw$n_wells == 1]))`;
  the largest effect overall is `r sprintf("%.0fx", 2^max(top25_raw$effect))`, on
  `r top25_raw$n_wells[which.max(top25_raw$effect)]` wells.
- **DIV 5 is excluded entirely.** Values are a fraction of same-plate vehicle controls, and
  at DIV 5 those controls are themselves silent: the control mean is exactly zero for
  network spikes in **5 of 6 experiments**, leaving the ratio undefined for **63–81% of
  wells on the network metrics** (0% on firing rate, active electrodes and synchrony).
  Cortical networks have not formed at DIV 5; that is what this assay measures.
- **No timepoint here is pre-treatment.** Compounds are applied from the start of the
  12-day culture, so "change across DIV" is not "onset of effect".
- **Plate, batch and dose structure are not modelled.** Wells nest in cultures, cultures
  reuse physical plates, doses nest in compounds, and DIV is a repeated measure on the
  same wells. Conditions are treated as flat groups.
- **Prediction is not causation.** Recovering the training set's own active/inactive labels
  says the assay predicts those labels. It says nothing about developmental neurotoxicity
  in humans.
- **The burst-interval and network-spike *shape* metrics are excluded from the state
  space** (`mean.isis`, `mean.dur`, `mean.IBIs`, `ns.peak.m`, `ns.durn.m`,
  `ns.mean.insis`, `ns.durn.sd`, `ns.mean.spikes.in.ns`), because their 33–42% missingness
  means "this well did not burst" rather than "measurement lost", and mean-imputing would
  invent burst statistics for silent wells. The network-spike **counts** (`ns.n`,
  `ns.percent.of.spikes.in.ns`) have no missingness and *are* in the state space.
:::

```{r provenance, results="asis"}
src <- file.path(params$study_dir, params$source_md)
cat("## Source\n\n")
if (knitr::is_html_output()) {
  cat(paste(readLines(src, warn = FALSE), collapse = "\n"), "\n")
} else {
  cat("US EPA (Brown et al. 2016, *Toxicol Sci* 154(1):126-139, doi:10.1093/toxsci/kfw147), ",
      "public domain (17 USC 105). URL, SHA-256 checksum, the compound designations quoted ",
      "from the abstract, and the full structural record are in `SOURCE.md`, reproduced in ",
      "the HTML notebook.\n", sep = "")
}
cat("\n---\n\nGenerated with [NOVA](https://github.com/atudoras/nova) ",
    as.character(utils::packageVersion("NOVA")),
    ". Discovery pass: `run_discovery.R`. Figures use `nova_theme()`/`nova_palette()`. ",
    "Every result shown was selected by effect-size rank from `findings.csv`; ",
    "none was hand-picked.\n", sep = "")
```
