---
title: "NOVA Dynamics: a dynamical-systems toolkit for neuronal network state analysis"
author: "NOVA"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    fig_width: 7
    fig_height: 4.6
vignette: >
  %\VignetteIndexEntry{NOVA Dynamics Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.align = "center", dpi = 110,
  warning = FALSE, message = FALSE
)
set.seed(20260622)
```

## 1. From "where is the network?" to "where is it going?"

The original NOVA answers a spatial question: it embeds high-dimensional
Multi-Electrode Array (MEA) recordings into an interpretable state space (PCA)
and shows *where* a neuronal network sits. `nova_dynamics` formalises the
**temporal** question that NOVA was already gesturing at with its trajectory
plots:

> Where is the network *moving*, what state is it *approaching*, and how
> *stable* is that state?

A culture's path through state space is a curve $x(t)\in\mathbb{R}^d$. From that
curve we can read trajectory **geometry** (speed, directness), **transitions**
between coarse-grained states, **similarity** between conditions, the
**dynamical regime** (stable / convergent / divergent / oscillatory /
transitional), and the **occupancy landscape**. A rule-based layer,
`nova_describe()`, turns those numbers into plain-language interpretation.

```{r load}
library(NOVA)
library(ggplot2)
has_patchwork <- requireNamespace("patchwork", quietly = TRUE)
```

## 2. Load the dataset (real if available, else simulated)

This vignette runs the **standard NOVA workflow** on the bundled MEA agonist
dataset when it is present. On a clean machine (e.g. CRAN or a CI runner) the raw
CSVs may not ship with the installed package, so we fall back to a designed
synthetic dataset with the *same structure* (treatments x wells x ordered
timepoints) that contains one trajectory of each dynamical regime. Either way the
downstream code is identical — `nova_dynamics` only needs PCA coordinates, a
timepoint column, and a grouping column.

```{r load-data}
# --- Path A: the real workflow (run when the bundled data is found) ----------
example_dir <- system.file("extdata", "MEA Neuronal Agonists", package = "NOVA")
if (!nzchar(example_dir)) {
  alt <- file.path("..", "Example", "MEA Neuronal Agonists")  # building from source tree
  if (dir.exists(alt)) example_dir <- alt
}

pca <- NULL
if (nzchar(example_dir) && dir.exists(example_dir)) {
  proc <- tryCatch(
    process_mea_flexible(
      main_dir             = example_dir,
      selected_experiments = c("MEA012", "MEA013"),   # public WT plates
      grouping_variables   = c("Experiment", "Treatment", "Well"),
      baseline_timepoint   = "baseline",
      verbose              = FALSE),
    error = function(e) NULL)
  if (!is.null(proc)) {
    pca <- tryCatch(
      pca_analysis_enhanced(processing_result = proc,
                            grouping_variables = c("Treatment"), verbose = FALSE),
      error = function(e) NULL)
  }
}

data_source <- if (!is.null(pca)) "real MEA recordings (MEA012 + MEA013, WT)" else "simulated"
```

```{r simulate, eval = is.null(pca)}
# --- Path B: designed synthetic state space (one trajectory per regime) -------
# Timepoints deliberately mix units to exercise the time parser; baseline first.
timepoints <- c("baseline", "0min", "15min", "30min", "45min", "1h", "1h30min", "2h")
tmin <- nova_time_to_minutes(timepoints); tmin[1] <- -10           # baseline before 0
tnorm <- (tmin - min(tmin)) / (max(tmin) - min(tmin))              # 0..1 progress

# Each treatment is a prototype regime in PC space; wells add biological noise.
proto <- list(
  PBS_vehicle = function(s) cbind(0.15 * sin(2 * pi * s) * 0,    0 * s),          # stable
  KA_convergent = function(s) cbind(6 * (1 - exp(-3 * s)),       2 * (1 - exp(-3 * s))),
  Gabazine_divergent = function(s) cbind(-7 * s,                 5 * s^1.4),       # divergent
  DHPG_oscillatory = function(s) cbind(3 * sin(2.2 * pi * s),    3 * cos(2.2 * pi * s) - 3)
)
wells <- paste0("W", 1:6)
rows <- list()
for (tr in names(proto)) {
  base <- proto[[tr]](tnorm)
  for (w in wells) {
    jit <- matrix(rnorm(length(tnorm) * 2, 0, 0.45), ncol = 2)
    jit[1, ] <- 0                                                  # shared baseline
    xy <- base + jit
    rows[[paste(tr, w)]] <- data.frame(
      PC1 = xy[, 1], PC2 = xy[, 2], Treatment = tr, Well = w,
      Timepoint = timepoints, stringsAsFactors = FALSE)
  }
}
plot_data <- do.call(rbind, rows)
pca <- list(plot_data = plot_data,
            variance_explained = c(PC1 = 41.3, PC2 = 22.8))
```

```{r data-banner, echo = FALSE}
cat("Data source:", data_source, "\n")
cat("Treatments :", paste(unique(pca$plot_data$Treatment), collapse = ", "), "\n")
cat("Timepoints :", paste(nova_order_timepoints(pca$plot_data$Timepoint), collapse = " -> "), "\n")
```

## 3. Standard NOVA workflow: PCA and trajectories

`nova_dynamics` never replaces PCA — it consumes it. Here is the existing
trajectory view, with the **corrected timepoint ordering**: baseline is always
first and compound labels such as `1h15`/`1h30`/`1h45` sort by real elapsed time
rather than alphabetically.

```{r trajectories, fig.height = 4.4}
traj <- plot_pca_trajectories_general(
  pca, trajectory_grouping = "Treatment",
  timepoint_order = nova_order_timepoints(pca$plot_data$Timepoint),
  color_by = "Treatment", save_plots = FALSE, verbose = FALSE)
traj$plots$combined_average
```

The new ordering helper is available on its own and underpins every dynamics
function:

```{r ordering-demo}
nova_order_timepoints(c("1h30", "baseline", "1h", "1h15", "2h", "15min"))
nova_time_to_minutes(c("baseline", "30s", "45min", "1h30", "DIV7"))
```

## 4. Extract ordered state-space trajectories

Every analysis builds on one tidy object: ordered, replicate-averaged paths.

```{r extract}
tr <- nova_extract_trajectories(pca, group_var = "Treatment")
head(tr)
```

Note `time_numeric`: baseline is placed one median-gap *before* the first real
timepoint, so velocities use true elapsed time.

## 5. State-space geometry

```{r geometry, fig.height = 4.2}
geo <- nova_state_geometry(tr, verbose = FALSE)
geo$summary[, c("group", "path_length", "net_displacement",
                "straightness", "tortuosity", "directional_persistence")]
geo$plots$overlay
```

```{r geometry-velocity, fig.height = 4.6}
geo$plots$velocity
```

**Scientific reading.** *Path length* is total network reconfiguration;
*straightness* (net / path, in $[0,1]$) measures how directed the change was;
*directional persistence* (mean cosine between successive steps) separates
ballistic motion ($\approx 1$) from oscillation ($<0$). Velocity is colour-coded
along each segment, so an acute onset followed by adaptation is visible directly.

```{r geometry-desc, results = "asis"}
invisible(nova_describe(geo))
```

## 6. State-transition analysis

We coarse-grain the embedding into `k` network macrostates (k-means) and count
empirical state(t) -> state(t+1) transitions along each replicate path.

```{r transitions, fig.height = 4.4}
tm <- nova_transition_matrix(pca, group_var = "Treatment",
                             unit_var = "Well", k = 4, verbose = FALSE)
round(tm$transition, 2)
tm$plots$flow
```

```{r transitions-heatmap, fig.height = 4.0}
tm$plots$heatmap
```

The diagonal of $P$ is each state's persistence; occupancy (node size in the
flow diagram) shows where networks spend their time. With only a handful of
timepoints this is a **descriptive flow summary**, not a fitted Markov model — a
caveat `nova_describe()` states explicitly.

```{r transitions-desc, results = "asis"}
invisible(nova_describe(tm))
```

## 7. Trajectory similarity

Which treatments produce *similar paths*? We compare whole curves with Dynamic
Time Warping (shape, allowing local time-warping), or Fréchet / Euclidean /
cosine distances, then cluster.

```{r similarity, fig.height = 4.2}
sim <- nova_trajectory_similarity(pca, method = "dtw",
                                  group_var = "Treatment", n_clusters = 2,
                                  verbose = FALSE)
round(sim$distance, 2)
sim$plots$dendrogram
```

```{r similarity-desc, results = "asis"}
invisible(nova_describe(sim))
```

## 8. Dynamical regime detection

Each trajectory is scored against five canonical regimes from transparent,
tunable rules and assigned the best-fitting one with a confidence.

```{r regime, fig.height = 4.4}
reg <- nova_dynamical_regime(pca, group_var = "Treatment", verbose = FALSE)
reg$classification[, c("group", "regime", "confidence",
                       "path_norm", "straightness", "decel_ratio")]
reg$plots$overlay
```

```{r regime-scores, fig.height = 3.8}
reg$plots$scores
```

The six decision thresholds live in one editable block
(`.nova_regime_thresholds`) and can be overridden per call via the
`thresholds =` argument, so the classifier is fully auditable.

```{r regime-desc, results = "asis"}
invisible(nova_describe(reg))
```

## 9. State-occupancy landscape

Where do networks spend time? The kernel-density landscape highlights frequently
occupied regions; the pseudo-potential $U=-\log p$ turns dense regions into
attractor-like valleys (a descriptive heuristic, not a fitted energy).

```{r landscape, fig.height = 4.4}
lnd <- nova_landscape(pca, group_var = "Treatment", verbose = FALSE)
lnd$plots$density
```

```{r landscape-potential, fig.height = 4.4}
lnd$plots$potential
```

```{r landscape-desc, results = "asis"}
invisible(nova_describe(lnd))
```

## 10. One call, full narrative

`nova_dynamics()` runs the whole pipeline once; `nova_describe()` narrates it
end to end.

```{r pipeline}
dyn <- nova_dynamics(pca, group_var = "Treatment", unit_var = "Well",
                     verbose = FALSE)
dyn
```

```{r pipeline-desc, results = "asis"}
invisible(nova_describe(dyn))
```

```{r composite, eval = has_patchwork, fig.height = 7.5, fig.width = 8.5}
# A publication-style composite figure (requires the 'patchwork' package).
library(patchwork)
(dyn$geometry$plots$overlay | dyn$regime$plots$overlay) /
  (dyn$transitions$plots$flow | dyn$landscape$plots$potential) +
  patchwork::plot_annotation(
    title = "NOVA Dynamics overview",
    subtitle = paste("Data source:", data_source),
    theme = nova_theme())
```

## 11. Biological interpretation

In an acute neuronal-agonist experiment, the dynamical readout maps onto
pharmacology:

- A **vehicle** control should be **stable** — little net movement in state
  space — providing the null trajectory other conditions are measured against.
- An **excitatory agonist** that drives the network to a new firing regime and
  then plateaus reads as **convergent** (directed motion that decelerates toward
  a settling state).
- A **disinhibitory** or runaway-excitation condition that keeps pushing the
  network outward reads as **divergent**.
- A condition that cycles the network without net progress reads as
  **oscillatory**.

Because trajectory *shape* — not just endpoint — drives `nova_trajectory_similarity()`,
agonists with shared mechanism cluster together even when their magnitudes
differ, which is exactly the comparison drug-screening and genotype studies need.

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
