---
title: "NOVA: summarising how conditions move through state space"
author: "NOVA"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    fig_width: 7
    fig_height: 4.6
vignette: >
  %\VignetteIndexEntry{NOVA Trajectory Summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.align = "center", dpi = 110,
  warning = FALSE, message = FALSE
)
set.seed(20260622)
```

## 1. The idea

After PCA, each MEA culture is a point in state space, and a timecourse is a path
through it. NOVA's `nova_trajectory_summary()` describes that path **simply and
honestly**: how far each condition moved away from baseline, how directly, and
when — using the replicate wells to put error bands on the key figure.

It deliberately stops there. A typical MEA experiment has only a handful of
timepoints and a few wells per condition: enough to say *"this drug moved the
network far and fast and it held"*, but **not** enough to fit velocities,
"stable vs unstable" regimes, or transition models without turning noise into
false precision. So those are intentionally out of scope.

```{r load}
library(NOVA)
library(ggplot2)
```

## 2. Get a PCA result (real data if available, else simulated)

This runs the standard NOVA workflow on the bundled MEA agonist data when it is
present, and otherwise falls back to a small simulated dataset with the same
shape, so the vignette always builds.

```{r data}
example_dir <- system.file("extdata", "MEA Neuronal Agonists", package = "NOVA")
if (!nzchar(example_dir)) {
  alt <- file.path("..", "Example", "MEA Neuronal Agonists")
  if (dir.exists(alt)) example_dir <- alt
}

pca <- NULL
if (nzchar(example_dir) && dir.exists(example_dir)) {
  proc <- tryCatch(process_mea_flexible(
    main_dir = example_dir, selected_experiments = c("MEA012", "MEA013"),
    grouping_variables = c("Experiment", "Treatment", "Well"),
    baseline_timepoint = "baseline", verbose = FALSE), error = function(e) NULL)
  if (!is.null(proc))
    pca <- tryCatch(pca_analysis_enhanced(processing_result = proc,
      grouping_variables = c("Treatment", "Well"), verbose = FALSE), error = function(e) NULL)
}

if (is.null(pca)) {
  timepoints <- c("baseline", "0min", "15min", "30min", "1h", "1h30", "2h")
  tnum <- nova_time_to_minutes(timepoints); tnum[1] <- -10
  s <- (tnum - min(tnum)) / (max(tnum) - min(tnum))
  proto <- list(PBS = function(s) cbind(0 * s, 0 * s),
                KA  = function(s) cbind(5 * (1 - exp(-3 * s)), 2 * (1 - exp(-3 * s))),
                Gabazine = function(s) cbind(-6 * s, 4 * s))
  rows <- list()
  for (tr in names(proto)) for (w in paste0("W", 1:6)) {
    xy <- proto[[tr]](s) + matrix(rnorm(length(s) * 2, 0, 0.4), ncol = 2)
    xy[1, ] <- xy[1, ] * 0
    rows[[paste(tr, w)]] <- data.frame(PC1 = xy[, 1], PC2 = xy[, 2],
      Treatment = tr, Well = w, Timepoint = timepoints)
  }
  pca <- list(plot_data = do.call(rbind, rows),
              variance_explained = c(PC1 = 40, PC2 = 22))
}
cat("Timepoints:", paste(nova_order_timepoints(pca$plot_data$Timepoint), collapse = " -> "), "\n")
```

## 3. Trajectories (standard NOVA)

```{r traj, fig.height = 4.4}
plot_pca_trajectories_general(
  pca, trajectory_grouping = "Treatment",
  timepoint_order = nova_order_timepoints(pca$plot_data$Timepoint),
  color_by = "Treatment", save_plots = FALSE, verbose = FALSE)$plots$combined_average
```

## 4. Trajectory summary

```{r summary}
s <- nova_trajectory_summary(pca, group_var = "Treatment", verbose = FALSE)
s$metrics
```

The key figure — distance from baseline over time, with a mean ± SEM band across
replicate wells:

```{r disp, fig.height = 4.6}
s$plots$displacement
```

The same paths drawn in PC space:

```{r map, fig.height = 4.6}
s$plots$map
```

## 5. Plain-language summary

```{r describe, results = "asis"}
invisible(nova_describe(s))
```

## 6. How to read it

- **`net_displacement`** — how far the condition ended up from baseline.
- **`directness`** (`net / path`, 0–1) — close to 1 means the network moved
  straight out and stayed; low values mean it wandered or returned.
- **`peak_timepoint`** — when the displacement was largest, a simple readout of
  response kinetics (fast vs slow) without claiming a velocity.

In an acute neuronal-agonist experiment this lines up with the biology: a vehicle
control barely leaves baseline, while an excitatory agonist moves the network far
and early and then holds — exactly the comparison a drug or genotype screen needs,
described without over-interpreting the dynamics.

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
