---
title: "`r params$title`"
subtitle: "`r params$subtitle`"
date: today
params:
  title: "NOVA case study"
  subtitle: "Exploratory reanalysis of a public MEA dataset"
  study_dir: "."
  nova_root: ".."
  findings_csv: "outputs/discovery/findings.csv"
  data_rds: "outputs/discovery/discovery.rds"
  source_md: "SOURCE.md"
  group_var: "Treatment"
  n_top: 6
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    theme: cosmo
    fig-width: 8
    fig-height: 5
  typst:
    papersize: a4
    margin: {x: 2cm, y: 2cm}
    fig-format: png
    fig-dpi: 200
    fig-width: 6.5
    fig-height: 3.6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Stage 2: the curated, portfolio-facing summary.
#
# It selects nothing by hand. Everything below is chosen by effect-size rank from
# the stage-1 findings table, so what is shown is traceable to `findings.csv`
# rather than to anyone's eye for a nice result.
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2); library(knitr)
})

# Load NOVA from this repo's source, never from the installed library. The case
# studies live inside the package repo and exist to exercise its current code;
# an installed copy can silently be an older version with different behaviour
# (and different numbers), which is exactly what a portfolio must not do.
if (!requireNamespace("devtools", quietly = TRUE)) {
  stop("devtools is required to load NOVA from source for this report.")
}
suppressMessages(devtools::load_all(params$nova_root, quiet = TRUE))
stopifnot("nova_trajectory_summary" %in% getNamespaceExports("NOVA"))

findings <- read.csv(file.path(params$study_dir, params$findings_csv),
                     stringsAsFactors = FALSE)
scan <- if (file.exists(file.path(params$study_dir, params$data_rds))) {
  readRDS(file.path(params$study_dir, params$data_rds))
} else NULL

moves <- findings[findings$kind == "move", ]
traj  <- findings[findings$kind == "trajectory_distance", ]
sep   <- findings[findings$kind == "group_separation", ]
load  <- findings[findings$kind == "loading", ]

top_moves <- head(moves, params$n_top)
n_scanned <- nrow(moves)
```

## Headline

```{r headline, results="asis"}
if (nrow(traj)) {
  far <- traj[which.max(traj$effect), ]
  near <- traj[which.min(traj$effect), ]
  cat(sprintf(
    paste("Across %d conditions, **%s** moved farthest from baseline in state space",
          "(%.2f PC units) and **%s** moved least (%.2f).",
          "The largest single baseline-normalised change was **%s** in **%s** at **%s**",
          "(%.1f-fold %s).\n"),
    nrow(traj), far$group, far$effect, near$group, near$effect,
    if (nrow(top_moves)) top_moves$metric[1] else "n/a",
    if (nrow(top_moves)) top_moves$group[1] else "n/a",
    if (nrow(top_moves)) top_moves$timepoint[1] else "n/a",
    if (nrow(top_moves)) 2^top_moves$effect[1] else NA_real_,
    if (nrow(top_moves)) top_moves$direction[1] else ""))
}
```

## Hero figure — distance from baseline

```{r hero, fig.cap="Distance from baseline in PC space, mean +/- SEM across replicate wells. Each well is identified by plate and well ID, so the same well ID on two plates counts as two replicates."}
if (!is.null(scan$trajectory)) {
  scan$trajectory$plots$displacement
} else {
  cat("Discovery result not found; rerun the discovery pass.")
}
```

::: {.content-visible when-format="html"}

## Where the variance sits

```{r pca, fig.cap="PC1/PC2 scores coloured by condition. PCA is on baseline-normalised values, scaled and centred; scaling makes metrics on different units comparable but also lets low-variance metrics contribute equally."}
if (!is.null(scan$pca)) {
  pd <- scan$pca$plot_data
  gv <- params$group_var
  ve <- scan$pca$variance_explained
  ggplot(pd, aes(.data[["PC1"]], .data[["PC2"]], colour = .data[[gv]])) +
    stat_ellipse(type = "norm", level = 0.95, linewidth = 0.7, alpha = 0.6) +
    geom_point(size = 2, alpha = 0.55) +
    scale_colour_manual(values = NOVA::nova_palette(dplyr::n_distinct(pd[[gv]])),
                        name = gv) +
    labs(x = sprintf("PC1 (%.1f%%)", ve[["PC1"]]),
         y = sprintf("PC2 (%.1f%%)", ve[["PC2"]]),
         caption = "95% confidence ellipses") +
    NOVA::nova_theme()
}
```

:::

## Small multiples — the largest moves

```{r small-multiples, fig.height=5, fig.cap="The top-ranked baseline-normalised changes, selected by effect size from findings.csv. Dashed line = baseline (no change). Bars are median log2 fold-change; the well count for each is printed, because a large effect measured on few wells is not the same result as a large effect measured on many."}
if (nrow(top_moves)) {
  d <- top_moves
  d$label <- paste0(d$metric, "\n", d$group, " @ ", d$timepoint)
  d$signed <- ifelse(d$direction == "down", -d$effect, d$effect)
  d$label <- factor(d$label, levels = d$label[order(d$signed)])
  ggplot(d, aes(x = .data$label, y = .data$signed, fill = .data$group)) +
    geom_col(width = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text(aes(label = paste0("n=", .data$n_wells)),
              hjust = ifelse(d$signed >= 0, -0.2, 1.2), size = 3, colour = "grey30") +
    scale_fill_manual(values = NOVA::nova_palette(dplyr::n_distinct(d$group)),
                      name = params$group_var) +
    coord_flip() +
    labs(x = NULL, y = "median log2 fold-change vs baseline") +
    NOVA::nova_theme(base_size = 10)
}
```

## In plain language

```{r describe, results="asis"}
# nova_describe() is rule-based, not generative: it restates the trajectory
# metrics without inferring mechanism.
if (!is.null(scan$trajectory)) {
  txt <- utils::capture.output(NOVA::nova_describe(scan$trajectory))
  cat("```\n"); cat(txt, sep = "\n"); cat("\n```\n")
}
```

::: {.content-visible when-format="html"}

```{r ranked-table}
if (nrow(top_moves)) {
  top_moves %>%
    transmute(Condition = .data$group, Metric = .data$metric,
              Timepoint = .data$timepoint,
              `Fold-change` = sprintf("%.2fx %s", 2^.data$effect, .data$direction),
              Wells = .data$n_wells, Plates = .data$n_plates) %>%
    knitr::kable(caption = "Top baseline-normalised moves, by effect size.")
}
```

:::

## Caveats

::: {.callout-warning appearance="simple"}
**These findings are exploratory. None of them is a tested claim.**

- **Selection inflates the top of the list.** `r n_scanned` condition x metric x
  timepoint comparisons were scanned and sorted by effect size. Sorting a large
  family of noisy estimates puts the luckiest at the top, so the effects shown here
  are biased upward and would be expected to shrink on replication. No multiple-
  comparison correction is applied, because nothing here is a hypothesis test.
- **Ranking is by effect size, not by evidence.** A large fold-change measured on
  three wells outranks a smaller one measured on thirty. The well and plate counts
  are printed for exactly this reason — read them.
- **Fold-changes are ratios**, so they are asymmetric and undefined against a zero
  baseline. Wells whose baseline is near zero are excluded from the ranking rather
  than allowed to produce enormous artefactual effects.
- **Plate and batch structure is not modelled.** Wells are nested in plates and
  timepoints are repeated measures on the same wells; both violate the independence
  a naive test would assume.
- **Prediction is not causation.** That a metric separates conditions does not mean
  it drives the biology, and a condition moving in state space describes what was
  measured, not a mechanism.
- **PCA choices are consequential**: values are scaled and centred, so every metric
  contributes equally regardless of how much it actually varies.
:::

```{r provenance, results="asis"}
# The PDF brief carries a pointer; the HTML notebook carries the full record.
src <- file.path(params$study_dir, params$source_md)
cat("## Source\n\n")
if (!file.exists(src)) {
  cat("_No SOURCE.md recorded for this dataset._\n")
} else if (knitr::is_html_output()) {
  cat(paste(readLines(src, warn = FALSE), collapse = "\n"), "\n")
} else {
  cat("Full provenance — URL, licence, download date and checksum — is recorded in ",
      "`", params$source_md, "` alongside this study, and reproduced in the HTML ",
      "notebook.\n", sep = "")
}
cat("\n\n---\n\nGenerated with [NOVA](https://github.com/atudoras/nova) ",
    as.character(utils::packageVersion("NOVA")),
    ". Figures use `nova_theme()` / `nova_palette()`. ",
    "Curated automatically by effect-size rank from `findings.csv` — ",
    "no result was hand-picked.\n", sep = "")
```
