---
title: "Can a generic pipeline pick out the neurotoxins?"
subtitle: "NOVA vs. the US EPA's developmental-neurotoxicity screen"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
suppressPackageStartupMessages({ library(dplyr); library(ggplot2) })
suppressMessages(devtools::load_all("../..", quiet = TRUE))   # NOVA from source
```

## The data

The EPA screened six compounds for developmental neurotoxicity by growing rat cortical
networks on 48-well Axion MEAs, dosing them from day one, and recording as the networks
formed — DIV 5, 7, 9 and 12. Six cultures, 256 wells, nine concentrations per compound
plus vehicle. Public domain, and openly downloadable.

> Brown JP, Hall DL, Frank CL, Wallace K, Mundy WR, Shafer TJ (2016). *Evaluation of a
> Microelectrode Array-Based Assay for Neural Network Ontogeny Using Training Set
> Chemicals.* **Toxicological Sciences** 154(1):126–139.
> [doi:10.1093/toxsci/kfw147](https://doi.org/10.1093/toxsci/kfw147) ·
> [dataset](https://catalog.data.gov/dataset/data-for-brown-et-al-mea-developmental-neurotoxicity-screening-manuscript)

Their abstract states the answer key, which is what makes this a fair test:

> "All compounds except acetaminophen (≤ 30 µM) caused concentration-related effects on
> one or more of these parameters."

Five actives, one negative. **NOVA is told none of it.**

```{r}
#| label: download
url <- paste0("https://pasteur.epa.gov/uploads/10.23719/1412682/",
              "Final_Data_Set_SA1_DNT_Paper1%20%282%29%28updated%29_CF.csv")
csv <- "data/brown2016_mea_dnt.csv"

if (!file.exists(csv)) {
  dir.create("data", showWarnings = FALSE)
  download.file(url, csv, quiet = TRUE)
}
# The file this notebook was written against, so a silent upstream change is caught.
stopifnot(digest::digest(file = csv, algo = "sha256") ==
          "80522f6654012a18851ad17f180f51d98df6b84ab122078ce303c4d3c079f43a")

raw <- read.csv(csv, stringsAsFactors = FALSE)
dim(raw)
```

## The analysis

This is the whole thing.

```{r}
#| label: process
# The eight metrics with no missing values. The rest are only defined for wells
# that actually burst, and "did not burst" is not a missing measurement.
metrics <- c("meanfiringrate", "burst.per.min", "per.spikes.in.burst", "nAE",
             "nABE", "ns.n", "ns.percent.of.spikes.in.ns", "r")

d <- process_mea_table(
  raw[, c("date", "Plate.SN", "DIV", "well", "trt", "dose", metrics)],
  # Plate serials are REUSED across culture dates -- 4 serials, 6 experiments.
  # Keying on the serial alone would merge two different cultures into one well.
  experiment = c("date", "Plate.SN"),
  well       = "well",
  timepoint  = "DIV",
  treatment  = "trt",
  metrics    = metrics,
  timepoint_prefix = "DIV",
  # Each well as a fraction of the vehicle wells on its own plate, at its own DIV.
  # (Not vs. DIV5: at DIV5 the controls are themselves silent, so the ratio is
  # undefined for most wells -- the networks have not formed yet. That is the
  # phenomenon, not a defect.)
  normalize  = "control",
  control    = raw$dose == 0,
  verbose    = FALSE
)
```

Rank every compound × dose × metric × DIV by how far it moved from its controls. No test,
no model — just the size of the change, on conditions with at least four wells.

```{r}
#| label: rank
nd <- d$normalized_data
nd$unit <- nova_unit_id(nd)   # a well is its plate AND its ID: "A1" is on every plate

moves <- nd |>
  filter(Timepoint != "DIV5", !Is_Control,
         is.finite(Normalized_Value), Normalized_Value > 0) |>
  group_by(compound = Treatment, dose, Variable, Timepoint) |>
  summarise(log2fc = median(log2(Normalized_Value)),
            wells  = n_distinct(unit), .groups = "drop") |>
  filter(wells >= 4) |>
  arrange(desc(abs(log2fc)))

head(moves, 5)
```

## What NOVA found

```{r}
#| label: scorecard
#| echo: false
actives  <- c("Domoic Acid", "Sodium Orthovanadate", "Bisindolymaleimide 1",
              "Mevastatin", "Loperamide")
top <- head(moves, 25)
knitr::kable(tibble::tibble(
  ` ` = c("Among the paper's five actives", "Acetaminophen (the negative)", "Decreases"),
  `of the top 25` = c(sum(top$compound %in% actives),
                      sum(top$compound == "Acetaminophen"),
                      sum(top$log2fc < 0))
))
```

**Every one of the 25 largest changes is one of the five compounds the paper calls
active. The designated negative never appears.**

```{r}
#| label: hero
#| echo: false
#| fig-width: 7.5
#| fig-height: 4
sync <- moves |>
  filter(Variable == "r", Timepoint == "DIV7") |>
  mutate(fc = 2^log2fc,
         Designation = ifelse(compound == "Acetaminophen", "negative", "active"))

ggplot(sync, aes(dose, fc, colour = compound)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_line(aes(linetype = Designation), linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = nova_palette(n_distinct(sync$compound)), name = NULL) +
  scale_linetype_manual(values = c(active = "solid", negative = "22"), name = NULL) +
  labs(title = "Network synchrony as a fraction of vehicle controls, DIV 7",
       x = "Dose (µM, log)", y = "Synchrony / vehicle (log)") +
  nova_theme(base_size = 11)
```

Sodium orthovanadate takes network synchrony to **1% of vehicle**, in dose order.
Acetaminophen is flat across three orders of magnitude. Nobody told NOVA which was which.

## The catch

`r nrow(moves)` condition × metric × DIV comparisons were scanned and sorted by effect
size. Sorting a large family of noisy estimates puts the luckiest on top, so these effects
are inflated and would shrink on replication. Nothing here is a hypothesis test.

More importantly: `r sum(top$compound %in% c("Bisindolymaleimide 1", "Mevastatin", "Loperamide"))`
of those top 25 are compounds the same paper reports as **cytotoxic**. The viability data
that would separate "changed the network" from "killed the cells" is not in this file. The
paper singles out domoic acid and sodium orthovanadate as the two that acted *without*
cytotoxicity — so "NOVA recovers the actives" is partly "NOVA recovers cell death".

This recovers labels the paper had already assigned. That is a claim about **prediction**,
not about neurotoxicity in humans.

<sub>Wells are nested in cultures, cultures reuse plates, doses nest in compounds, and DIV
is a repeated measure — none of it modelled. Fully-silenced wells are absent from the
ranking, because a log-ratio cannot represent zero.</sub>
